---
title: "Down and Out"
author: "Eric Green"
date: "September 4, 2014"
output: pdf_document
---

<!--- this is a hidden comment -->
<!--- Let's start with a little file setup
Ready....
-->

```{r, include=FALSE}
# this is an R code chunk
# remember, R can't read comments following a hash in this section
# this chunk won't print to the pdf because we said include=FALSE
  library(knitr)
  opts_chunk$set(message=FALSE)          # turn off messages
  options(scipen=999)                    # turn off scientific notation
  knit_hooks$set(inline = function(x) {  # thousands sep
    prettyNum(x, big.mark=",")
  })
```

This is my fancy report for class. I am going to show you how to use the `WDI` package to use data from the World Bank. The examples come from Vincent Arel-Bundock's [GitHub page](https://github.com/vincentarelbundock/WDI).

Let's search for "gdp" in the indicator names. There are a bunch, so I will just show the first 5. 

```{r, results='markup'}
# load packages
  library(WDI)
  library(ggplot2)
  WDIsearch('gdp')[1:5,]
```

Next I will download and plot some data.

```{r}
# download data
  dat <- WDI(indicator='NY.GDP.PCAP.KD', 
             country=c('MX','CA','US'), 
             start=1960, end=2012)
#   this creates an object called dat
# plot data
  ggplot(dat, 
         aes(year, NY.GDP.PCAP.KD, color=country)) + 
         geom_line() + 
         xlab('Year') + 
         ylab('GDP per capita')
  us.gdppc.12 <- dat$NY.GDP.PCAP.KD[dat$country=="United States" & 
                                    dat$year=="2012"]
```

GDP per capita for the US in 2012 was USD `r us.gdppc.12`. I know that `r us.gdppc.12` divided by 2 is `r us.gdppc.12/2`.